New cardinality estimation algorithms for HyperLogLog sketches

### 2016-11-19 paper/paper.tex
some more imporvements

In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available. On the one hand there is the self-learning bitmap \cite{Chen2011, Chen2015}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}. Both achieve the same estimation error using less space. However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.

@@

\begin{table}
\centering
\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated. The last column shows the average number of needed log-likelihood function 
 \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} until convergence of the BFGS algorithm.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{1}{c}{avg number}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
\multicolumn{1}{c}{iterations}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumEvaluations=\avgNumEvaluations
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumEvaluations}
}
\end{table}

@@

Using the maximum-likelihood method we have been able to improve the cardinality estimates of intersections and relative complements of two HyperLogLog sketches. Unfortunately, finding the maximum using the BFGS algorithm requires many iterations which makes the joint cardinality estimation much more expensive than for a single HyperLogLog sketch. Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case. It is expected that such a new algorithm makes use of all the information given by the sufficient statistic.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\left(
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetASuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
+
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
\right).
\end{equation}
Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
\symAlpha_\infty
\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
+
(1-\symAlpha_\infty).
\end{equation}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash functions evaluations are necessary for all the needed permutations \cite{Li2011}.

>>>>

In case the data is not distributed and results do not need to be aggregated further, there are even more efficient estimation algorithms available. On the one hand there is the self-learning bitmap \cite{Chen2011}, and on the other hand there is the HyperLogLog algorithm extended by a historic inverse probability estimator that is continuously updated together with the HyperLogLog sketch \cite{Ting2014}. Both achieve the same estimation error using less space. However, the estimated cardinality depends on the insertion order of elements and hence cannot be used in a distributed environment.

@@

\begin{table}
\centering
\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated. The last column shows the average number of log-likelihood function 
 \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi} evaluations until the stop criterion of the BFGS algorithm was satisfied.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{1}{c}{avg number}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
\multicolumn{1}{c}{iterations}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumEvaluations=\avgNumEvaluations
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumEvaluations}
}
\end{table}

@@

Using the maximum-likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches. Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function. Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case. It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\left(
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetASuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
+
\log\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
\right).
\end{equation}
Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
\symAlpha_\infty
\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
+
(1-\symAlpha_\infty).
\end{equation}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.
